Approximating the Number of COVID-19 Infections in the United States Across 2021

### 2023-07-24 reports/county_figures.Rmd
cross correlation figure

>>>>

# Cross Correlations

```{r plot ccf functions}

objs <- ls()
remove <- objs[which(!objs %in% c("theme_c", "wjoined", "pal"))]
rm(remove)

pal <- c("red", "#10BAC5", "#1B10C5", "#EFB719", "#900C3F", "#FF7400")

names(pal) <- c( "Covidestim",
                '$Pr(S_1|untested)$ Centered at Survey Value',
                '$Pr(S_1|untested)$ and $\\beta$ Centered at Survey Values',
               "$Priors\\,Do\\,Not\\,Vary\\,by\\,State\\,and\\,Date$",
               "$\\beta$ Centered at Survey Value",
               "observed"
               )

plot_all_ccf <- function(input_df_for_fips, 
                         varnames= c("differenced_mean_conc",
                                  "differenced_exp_median")) {

  versions <- input_df_for_fips %>% 
    group_split(version)

  message(length(versions))

  plt1 <- plot_ccf(versions[[1]], varnames)
  plt2 <- plot_ccf(versions[[2]], varnames)
  plt3 <- plot_ccf(versions[[3]], varnames)
  plt4 <- plot_ccf(versions[[4]], varnames)

  cowplot::plot_grid(plt1, plt2, plt3, plt4, nrow = 1)

plot_ccf <- function(input_df_for_version,
                     varnames = c("differenced_mean_conc",
                                  "differenced_exp_median")) {

  var1 <- varnames[1]
  var2 <-varnames[2]
  lab <- unique(input_df_for_version$vlabel)

  cross_correlations <- ccf(
    pull(input_df_for_version, var1),
     pull(input_df_for_version, var2), 
                            plot = FALSE)

  ccf_res <- tibble(lag = cross_correlations$lag[,,1], 
                  correlation=cross_correlations$acf[,,1])

  county_name <- gsub("County, MA", "", unique(input_df_for_version$name))

  ccf_res %>%  
    mutate(max = ifelse(correlation == max(correlation), 
                        lag, NA),
           maxcorr = round(max(correlation),3)) %>%
    ggplot(aes(x=lag, y = correlation)) +
    geom_point() +
    geom_linerange(aes(ymin = 0, ymax=correlation)) +
    theme_c() +
    theme(plot.title = element_text(size = 12, hjust = .5,
                                    margin=margin(1,1,8,1)),
          plot.margin = margin(10,1,1,1),
            plot.subtitle = element_text(size = 12, hjust = .5, 
                                         margin = margin(4,1,1,1))) +
    labs(title =TeX(paste0(
      "Cross Correlation between $(Y_t)$ and $(Z_t)$, County: ", 
      county_name)),
         subtitle = TeX(paste0("Version: ", lab))) +
    geom_label(aes(x = max, y = correlation,
                   label = paste0("Max Correlation: ",
                                  maxcorr, "\n Lag: ",
                                  max)),
               hjust = -.1) +
    ylim(-1,1) +
    scale_x_continuous(n.breaks = 10)

get_max_lag_by_version <- function(input_df_for_version,
                                   varnames = c("differenced_mean_conc",
                                  "differenced_exp_median"),
                                  comparison = "bias corrected") {

  var1 <- varnames[1]
  var2 <-varnames[2]
  lab <- unique(input_df_for_version$vlabel)

  cross_correlations <- ccf(
    pull(input_df_for_version, var1),
     pull(input_df_for_version, var2), 
                            plot = FALSE)

  # cross_correlations <- ccf(input_df_for_version$differenced_mean_conc,
  #                           input_df_for_version$differenced_exp_median, 
  #                           plot = FALSE)

  ccf_res <- tibble(lag = cross_correlations$lag[,,1], 
                  correlation=cross_correlations$acf[,,1])

  m <- ccf_res %>% filter(correlation == max(correlation))

  tibble(vlabel = unique(input_df_for_version$vlabel),
         fips = unique(input_df_for_version$fips),
         max_lag = m$lag,
         max_corr = m$correlation,
         comparison = comparison,
         county_name = unique(input_df_for_version$name))

```{r generate differenced df}

differenced_all <- wjoined %>%
  filter(!is.na(mean_conc)) %>%
  select(-date) %>%
  distinct() %>%
  group_by(vlabel, fips) %>%
  arrange(biweek) %>%
  mutate(differenced_exp_median = exp_cases_median -
           lag(exp_cases_median, n =1),
         differenced_mean_conc = mean_conc - lag(mean_conc, n = 1),
         differenced_positive = positive - lag(positive, n = 1),
         differenced_covidestim = infections - lag(infections, n = 1)) %>%
  ungroup() %>%
  filter(!is.na(differenced_exp_median) & !is.na(differenced_mean_conc)) %>%
  ungroup()

### Cross Correlation (Wastewater and Covidestim Estimates)

```{r wastewater covidestim, fig.width = 16}

# glimpse(differenced_all)

differenced_all %>% 
  filter(version == "v7" & !is.na(differenced_covidestim)) %>%
  group_by(fips) %>%
  group_split() %>% 
  walk(~ { 
    plt <- .x %>% plot_ccf(varnames= c("differenced_mean_conc",
                                  "differenced_covidestim"))
    print(plt)})

### Cross Correlation (Wastewater and Observed Cases)

```{r, error = FALSE, out.width = "80%"}

# cross correlation plots for each county
differenced_all %>% 
  filter(version == "v7") %>%
  group_by(fips) %>%
  group_split() %>% 
  walk(~ { 
    plt <- .x %>% plot_ccf(varnames= c("differenced_mean_conc",
                                  "differenced_positive"))
    print(plt)})

### Cross Correlation (Wastewater and Bias-Corrected Cases)

```{r, fig.width = 23}

# cross correlation plots for each county
differenced_all %>% 
  group_by(fips) %>%
  group_split() %>%
  walk(~ { 
    plt <- .x %>% plot_all_ccf()
    print(plt)})

# Compare Lags with Maximum Correlation Between Covidestim, Wastewater, and Bias Corrected Counts

## Compare Lags Among versions of Bias Corrected Counts

For each county, lag with the maximum correlation:

```{r, fig.width = 12.7, fig.height =7}

# lags for bias-corrected counts
lags_all_pb <- differenced_all %>%
  group_by(fips, vlabel) %>%
  group_split() %>%
  map_df(get_max_lag_by_version)

lags_all_pb %>%
  group_by(fips) %>%
  summarize(`Lags with Maximum Correlation in Some Version` =
              paste0(unique(max_lag), collapse =", "))

# lags for observed counts
lags_all_observed <- differenced_all %>% 
  filter(version == "v7") %>%
  group_by(fips) %>%
  group_split()%>%
  map_df(~get_max_lag_by_version(
    input_df_for_version = .x,
    varnames= c("differenced_mean_conc",
                "differenced_positive"),
   comparison = "observed" )) %>%
  select(-vlabel)

# compare pb max correlations to those for observed counts 
lags_all_pb %>%
  bind_rows(lags_all_observed) %>%
  mutate(vlabel=as.character(vlabel)) %>%
   mutate(source = ifelse(!is.na(vlabel), vlabel, comparison)) %>%
  group_by(fips) %>%
  mutate(m = median(max_corr)) %>%
  ungroup() %>%
  mutate(max_lag=factor(max_lag)) %>%
  mutate(county_name = gsub("County, MA", "", county_name)) %>%
   ggplot(aes(x=fct_reorder(county_name,m, 
                           .desc = TRUE),
             y = max_corr,
             color = source, 
             shape = max_lag)) +
  geom_jitter(aes(fill=source),color="black",
              height =0, width =.15, size = 3,
              stroke=.1,
              alpha=.9) +
  scale_shape_manual(values=c(21,22,24)) +
  scale_fill_manual(values=pal,
                    name = 'Time Series Compared with\nWastewater:',
                    labels=TeX(names(pal))) +
  theme_bw() +
  theme_c() +
  theme(plot.title = element_text(size=16,hjust=.2, face="bold"),
    # axis.text.x = element_text(angle  = 60, vjust = .4),
    axis.text.x = element_text(angle  = 50, vjust = .95, hjust=1, size =14),
    axis.title.x = element_text(margin = margin(3,3,3,3)),
        legend.text = element_text(size = 14, hjust=0),
        axis.text.y = element_text(size = 11),
        legend.title = element_text(face="bold", size = 16)) +
  labs(y = "Maximum Correlation") +
  scale_size_discrete(breaks = c(1.9,2),
                      range = c(1.9,2),
                      labels = c("bias corrected", "observed")) +
  labs(x = "",
       y = "Maximum Correlation",
       title = "Comparing Maximum Correlation of Probabilistic Bias Counts with Wastewater\nto that Between Observed Cases and Wastewater",
       shape = "Lag Where Maximum\nCorrelation Was Obtained:",
       color = "Time Series Compared with Wastewater:") +
  guides(shape = guide_legend(override.aes = list(size = 4, stroke=.4)),
         fill= guide_legend(override.aes = list(size = 3.5, 
                                                shape=24, 
                                                stroke=.8)))

ggsave(here("figures/cross_correlation_observed_pb.pdf"), 
       height = 8, width=12)

# max series by rank; use for following comparison
max_series_by_rank <- lags_all_pb %>%
  group_by(fips) %>%
  mutate(r = rank(max_corr),
            series = vlabel[which.max(r)]) %>%
  group_by(series) %>%
  summarize(n=n()) %>%
  filter(n==max(n)) %>%
  pull(series)

### 2023-07-24 writing/manuscript.Rmd
cross correlation figure

In the first three implementations, the prior distributions specified for the bias parameters vary by state and date. In particular, we use survey data from the COVID-19 Trends and Impact Survey [@salomon2021a] to inform the prior distributions for one to two of the bias parameters: the probability of having symptoms among the untested population ($\Pr(S_1|\text{untested})$), and the parameter $\beta$, which represents the ratio of the test positivity among members of the untested population who have no symptoms to the overall test positivity. More detail on these implementations is provided in the [Methods](#Methods) section. In another, more simple, implementation, which is the only implementation presented at the state-level, the prior distributions are informed by the distributions from the survey data across all dates and states, but the same prior distributions are used for all states and time intervals considered.

@@

```{r sim-intervals-color-county, fig.width=15,  fig.cap = "Simulation intervals for counties in Massachusetts, colored by the implementation of probabilistic bias analysis. Only the two implementations that were consistently the highest and lowest among the implementations are included for clarity. In the   In the first three implementations, the priors vary by state and date, while for the fourth, they are the same for all states and time intervals considered. For the first implementation, we center the distribution of $\\beta$ at the ratio of the screening test positivity to the overall test positivity from the survey, and we center the distribution of $\\Pr(S_1|\\text{untested})$ at the percentage of the population experiencing COVID-19-like illness from the survey for each two-week interval. The second implementation, we center only $\\beta$ at the aformentioned value, and the for the next we only center $\\Pr(S_1|\\text{untested})$ at the aforementioned value. The last implementation corresponds to the implementation where we specify priors that are the same for all dates. The implementation that centers both $\\Pr(S_1|\\text{untested})$ and $\\beta$ at the survey values is consistently the highest among the implementations, followed by the implementation that centers only $\\beta$ at the survey value, followed then by the implementation only centering $\\Pr(S_1|\\text{untested})$ at the survey value, and then the lowest among the implementations is that where the priors do not vary by date."}

@@

Also of interest beyond the total numbers of infections is the ratio of the infections that are estimated to those that were observed, as this gives us a sense for the case ascertainment at different time periods in the pandemic. We see in Figure \@ref(fig:ma-heatmap) that the two-week interval estimated to have the highest ratio of estimated to observed infections was July 2, 2021 through July 30, 2021, during the delta wave. This corresponds to a similar time frame that we saw at the state-level in Figure \@ref(fig:ratio-estimated-to-observed-by-wave). Hampsphire County and Suffolk County were consistently among the lowest with regard to the ratio of estimated to observed infections, which may be explained by the screening testing occurring at universities in these counties.

@@

```{r,eval=FALSE}

include_graphics(here('figures/simulation-intervals-county.pdf'))

>>>>

In the three out of the four implementations, the prior distributions specified for the bias parameters vary by state and date. In particular, we use survey data from the COVID-19 Trends and Impact Survey [@salomon2021a] to inform the prior distributions for one to two of the bias parameters: the probability of having symptoms among the untested population ($\Pr(S_1|\text{untested})$), and the parameter $\beta$, which represents the ratio of the test positivity among members of the untested population who have no symptoms to the overall test positivity. More detail on these implementations is provided in the [Methods](#Methods) section. In another, more simple, implementation, which is the only implementation presented at the state-level (Figure \@ref(fig:ratio-estimated-to-observed-by-wave) and \@ref(fig:simulation-intervals-state)), the prior distributions are informed by the distributions from the survey data across all dates and states, but the same prior distributions are used for all states and time intervals considered.

@@

```{r sim-intervals-color-county, fig.width=15,  fig.cap = "Simulation intervals for counties in Massachusetts, colored by the implementation of probabilistic bias analysis. Only the two implementations that were consistently the highest and lowest among the implementations are included for clarity. In the   In the three out of the four implementations, the priors vary by state and date, while for the fourth, they are the same for all states and time intervals considered. For the first implementation, we center the distribution of $\\beta$ at the ratio of the screening test positivity to the overall test positivity from the survey, and we center the distribution of $\\Pr(S_1|\\text{untested})$ at the percentage of the population experiencing COVID-19-like illness from the survey for each two-week interval. The second implementation, we center only $\\beta$ at the aformentioned value, and the for the next we only center $\\Pr(S_1|\\text{untested})$ at the aforementioned value. The last implementation corresponds to the implementation where we specify priors that are the same for all dates. The implementation that centers both $\\Pr(S_1|\\text{untested})$ and $\\beta$ at the survey values is consistently the highest among the implementations, followed by the implementation that centers only $\\beta$ at the survey value, followed then by the implementation only centering $\\Pr(S_1|\\text{untested})$ at the survey value, and then the lowest among the implementations is that where the priors do not vary by date."}

@@

Also of interest beyond the total numbers of infections is the ratio of the infections that are estimated to those that were observed, as this gives us a sense for the case ascertainment at different time periods in the pandemic. We see in Figure \@ref(fig:ma-heatmap) that the two-week interval estimated to have the highest ratio of estimated to observed infections was July 2, 2021 through July 30, 2021, during the delta wave. This corresponds to a similar time frame that we saw at the state-level in Figure \@ref(fig:ratio-estimated-to-observed-by-wave). Hampshire County and Suffolk County were consistently among the lowest with regard to the ratio of estimated to observed infections, which may be explained by the screening testing occurring at universities in these counties.

@@

Also of interest is whether there is a lag between the wastewater concentrations and bias corrected estimates. Although wastewater concentrations can be a leading indicator since people may shed viral material before developing symptoms and subsequently getting tested, the lead time is a result of multiple epidemiological factors, including viral shedding dynamics that may differ with evolution of the virus, access to testing, and testing behavior [@olesen2021a]. Indeed, the lead or lag time between wastewater concentrations and hospitalizations or cases has varies substantially between waves, where the lead time is higher in earlier waves and near zero at later waves, a difference that may be attributable to changes in diagnostic test availability over time  [@xiao2022a; @hopkins2023a].

In Figure \@ref(fig:cross-corr-wastewater), we see that among all counties except one in Massachsuetts, the maximum correlation between mean wastewater concentration and biweekly infection estimates was obtained at a lag of zero, that is, there  typically there was no lag among the biweekly infection estimates and mean wastewater concentrations. This is to be expected given the lead time for wastewater concentrations, when observed, is less than 2 weeks [@olesen2021a]. In two counties, Hamsphire and Barnstable, the lag between the probabilistic bias counts and wastewater concentration was less than that between the observed positive tests and wastewater concentrations.

```{r cross-corr-wastewater, out.width="70%"}

include_graphics(here("figures/cross_correlation_observed_pb.jpeg"))
